---
title: Advanced REDCapR Operations
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    # css: styles.css

vignette: >
  %\VignetteIndexEntry{Advanced REDCapR Operations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette covers the the less-typical uses of [REDCapR](https://github.com/OuhscBbmc/REDCapR) to interact with [REDCap](https://projectredcap.org/) through its API.

```{r set_options}
#| echo    = FALSE,
#| results = 'hide'

report_render_start_time <- Sys.time()

library(knitr)
library(magrittr)
suppressPackageStartupMessages(requireNamespace("kableExtra"))

opts_chunk$set(
  comment = "#>",
  tidy    = FALSE
)
```

Next Steps {#nextsteps .emphasized}
==================================================================

Set project-wide values
==================================================================

There is some information that is specific to a REDCap project, as opposed to an individual operation.  This includes the (1) uri of the server, and the (2) token for the user's project.  This is hosted on a machine used in REDCapR's public test suite, so you can run this example from any computer.  Unless tests are running.

*Other than PHI-free demos, we strongly suggest storing tokens securely and avoiding hard-coding them like below.  Our recommendation is to store tokens [in a database](https://ouhscbbmc.github.io/REDCapR/articles/SecurityDatabase.html).  If that is not feasible for your institution, consider storing them in a secured csv and retrieving with [`REDCapR::retrieve_credential_local()`](https://ouhscbbmc.github.io/REDCapR/reference/retrieve_credential.html).*

```{r project_values}
library(REDCapR) #Load the package into the current R session.
uri                   <- "https://redcap-dev-2.ouhsc.edu/redcap/api/"
token_simple          <- "9A068C425B1341D69E83064A2D273A70"
token_longitudinal    <- "DA6F2BB23146BD5A7EA3408C1A44A556"
```

Converting from tall/long to wide
==================================================================

*Disclaimer*: Occasionally we're asked for a longitudinal dataset to be converted from a "long/tall format" (where  typically each row is one observation for a participant) to a "wide format" (where each row is on participant).  Usually we advise against it.  Besides all the database benefits of a long structure, a wide structure restricts your options with the stat routine.  No modern longitudinal analysis procedures (*e.g.*, growth curve models or multilevel/hierarchical models) accept wide.  You're pretty much stuck with repeated measures anova, which is very inflexible for real-world medical-ish analyses.  It requires a patient to have a measurement at every time point; otherwise the anova excludes the patient entirely.

However we like going wide to produce visual tables for publications, and here's one way to do it in R. First retrieve the dataset from REDCap.

```{r retrieve-longitudinal}
#| results = 'hold'

library(magrittr)
suppressPackageStartupMessages(requireNamespace("dplyr"))
suppressPackageStartupMessages(requireNamespace("tidyr"))
events_to_retain  <- c("dose_1_arm_1", "visit_1_arm_1", "dose_2_arm_1", "visit_2_arm_1")

ds_long <- REDCapR::redcap_read_oneshot(redcap_uri = uri, token = token_longitudinal)$data
ds_long %>%
  dplyr::select(study_id, redcap_event_name, pmq1, pmq2, pmq3, pmq4)
```

When widening only one variable (*e.g.*, `pmq1`), the code's pretty simple:

```{r widen-simple}
#| results = 'hold'

ds_wide <-
  ds_long %>%
  dplyr::select(study_id, redcap_event_name, pmq1) %>%
  dplyr::filter(redcap_event_name %in% events_to_retain) %>%
  tidyr::pivot_wider(
    id_cols     = study_id,
    names_from  = redcap_event_name,
    values_from = pmq1
  )
ds_wide
```

In some scenarios, multiple variables (*e.g.*, `pmq1` - `pmq4`) can be widened in a single [`tidyr::pivot_wider()`](https://tidyr.tidyverse.org/reference/pivot_wider.html) operation.  This example contains the additional wrinkle that the REDCap event names "first_dose" and "first_visit" are renamed "dose_1" and "visit_1", which will help all the values be dose and visit values be proper numbers.

```{r widen-typical}
pattern <- "^(\\w+?)_arm_(\\d)$"
ds_wide <-
  ds_long %>%
  dplyr::select(study_id, redcap_event_name, pmq1, pmq2, pmq3, pmq4) %>%
  dplyr::mutate(
    event = sub(pattern, "\\1", redcap_event_name),
    event = dplyr::recode(event, "first_dose"="dose_1", "first_visit"="visit_1"),
    arm   = as.integer(sub(pattern, "\\2", redcap_event_name))
  ) %>%
  dplyr::select(study_id, event, arm, pmq1, pmq2, pmq3, pmq4) %>%
  dplyr::filter(!(event %in%
    c("enrollment", "final_visit", "deadline_to_return", "deadline_to_opt_ou")
  )) %>%
  tidyr::pivot_wider(
    id_cols     = c(study_id, arm),
    names_from  = event,
    values_from = c(pmq1, pmq2, pmq3, pmq4)
  )

ds_wide
```

However, in other widening scenarios, it can be easier to go even longer/taller (*e.g.*, `ds_eav`) before reversing direction and going wide.

```{r widen-long-first}
ds_eav <-
  ds_long %>%
  dplyr::select(study_id, redcap_event_name, pmq1, pmq2, pmq3, pmq4) %>%
  dplyr::mutate(
    event = sub(pattern, "\\1", redcap_event_name),
    event = dplyr::recode(event, "first_dose" = "dose_1", "first_visit" = "visit_1"),
    arm   = as.integer(sub(pattern, "\\2", redcap_event_name))
  ) %>%
  dplyr::select(study_id, event, arm, pmq1, pmq2, pmq3, pmq4) %>%
  tidyr::pivot_longer(
    cols      = c(pmq1, pmq2, pmq3, pmq4),
    names_to  = "key",
    values_to = "value"
  ) %>%
  # For old versions of tidyr that predate `pivot_wider()`:
  # tidyr::gather(key=key, value=value, pmq1, pmq2, pmq3, pmq4) %>%
  dplyr::filter(!(event %in% c(
    "enrollment", "final_visit", "deadline_to_return", "deadline_to_opt_ou")
  )) %>%
  dplyr::mutate( # Simulate correcting for mismatched names across arms:
    key = paste0(key, "_", event)
  ) %>%
  dplyr::select(-event)

# Show the first 10 rows of the EAV table.
ds_eav %>%
  head(10)

# Spread the EAV to wide.
ds_wide_2 <-
  ds_eav %>%
  tidyr::pivot_wider(
    id_cols     = c(study_id, arm),
    names_from  = key,
    values_from = value
  )
# For old versions of tidyr that predate `pivot_wider()`:
# tidyr::spread(key=key, value=value)
ds_wide_2
```

Lots of packages and documentation exist.  Our current preference is the [tidyverse approach](https://tidyr.tidyverse.org/articles/pivot.html) to pivoting, but the [data.table approach](https://www.r-bloggers.com/2019/03/creating-blazing-fast-pivot-tables-from-r-with-data-table-now-with-subtotals-using-grouping-sets/) is worth considering if you're comfortable with that package.  This [Stack Overflow post](https://stackoverflow.com/questions/10589693/convert-data-from-long-format-to-wide-format-with-multiple-measure-columns/) describes several ways.  We recommend against the [reshape](https://CRAN.R-project.org/package=reshape) and [reshape2](https://CRAN.R-project.org/package=reshape2) packages, because their developers have replaced them with the [tidyr](https://CRAN.R-project.org/package=tidyr) functions described above.

Query the Underlying MySQL Database
==================================================================

If you require a feature that is not available from your instance's API, first upgrade your institution's REDCap instance and see if the feature has been added recently.  Second, check if someone has released the desired API-like features as an [REDCap External Module](https://redcap.vanderbilt.edu/consortium/modules/).

Third, you may need to query the database underneath REDCap's web server.  The [Transfer Credentials](https://ouhscbbmc.github.io/REDCapR/articles/SecurityDatabase.html#transfer-credentials) section of the [Security Database Vignette](https://ouhscbbmc.github.io/REDCapR/articles/SecurityDatabase.html#transfer-credentials) provides a complete example of using R to query the MySQL database through odbc.

We find it's best to develop the query in [MySQL Workbench](https://www.mysql.com/products/workbench/), then copy the code to R (or alternatively, use [`OuhscMunge::execute_sql_file()`](https://ouhscbbmc.github.io/OuhscMunge/reference/execute_sql_file.html)).

Here is an example that retrieves the `first_submit_time`, which is helpful if you need a timestamp from surveys that were not marked as completed.  Replace '444' with your pid, and 1001 through 1003 with the desired events.

```sql
SELECT
  p.participant_id      as participant_survey_id
  ,r.record             as record_id
  ,p.event_id
  ,e.descrip            as event_name
  ,r.first_submit_time
  ,r.completion_time

  -- ,p.*
  -- ,r.*
FROM redcapv3.redcap_surveys_participants     as p
  left  join redcapv3.redcap_surveys_response as r on p.participant_id = r.participant_id
  left  join redcapv3.redcap_events_metadata  as e on p.event_id       = e.event_id
WHERE
  p.survey_id = 444
  and
  p.event_id in (
    1001, -- start of the year
    1002, -- mid term
    1003  -- end of year
  )
```

SSL Options
==================================================================

The official [cURL site](https://curl.se/) discusses the process of using SSL to verify the server being connected to.

Use the SSL cert file that come with the `openssl` package.

```{r}
cert_location <- system.file("cacert.pem", package = "openssl")
if (file.exists(cert_location)) {
  config_options         <- list(cainfo = cert_location)
  ds_different_cert_file <- redcap_read_oneshot(
    redcap_uri     = uri,
    token          = token_simple,
    config_options = config_options
  )$data
}
```

Force the connection to use SSL=3 (which is not preferred, and possibly insecure).
```{r}
config_options <- list(sslversion = 3)
ds_ssl_3 <- redcap_read_oneshot(
  redcap_uri     = uri,
  token          = token_simple,
  config_options = config_options
)$data

config_options <- list(ssl.verifypeer = FALSE)
ds_no_ssl <- redcap_read_oneshot(
  redcap_uri     = uri,
  token          = token_simple,
  config_options = config_options
)$data
```

Convert SPSS Output to REDCap data dictionary
==================================================================

The solution <https://stackoverflow.com/a/51013678/1082435> converts levels specified in SPSS output like

```
SEX       0 Male
          1 Female

LANGUAGE  1 English
          2 Spanish
          3 Other
          6 Unknown
```

to a dropdown choices in a REDCap data dictionary like

```csv
Variable Values
SEX      0, Male | 1, Female
LANGUAGE 1, English | 2, Spanish | 3, Other | 6, Unknown
```

Session Information
==================================================================

For the sake of documentation and reproducibility, the current report was rendered in the following environment.  Click the line below to expand.

<details>
  <summary>Environment <span class="glyphicon glyphicon-plus-sign"></span></summary>
```{r session-info, echo=FALSE}
if (requireNamespace("sessioninfo", quietly = TRUE)) {
  sessioninfo::session_info()
} else {
  sessionInfo()
}
```
</details>

```{r session-duration, echo=FALSE}
report_render_duration_in_seconds <- round(as.numeric(difftime(Sys.time(), report_render_start_time, units = "secs")))
```

Report rendered by `r Sys.info()["user"]` at `r strftime(Sys.time(), "%Y-%m-%d, %H:%M %z")` in `r report_render_duration_in_seconds` seconds.
