<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Basic REDCapR Operations}
-->

<!-- rmarkdown v1 -->

# Basic REDCapR Operations
This vignette covers the the basic functions exposed by the [`REDCapR`](https://github.com/OuhscCcanMiechvEvaluation/REDCapR) package which allow you to interact with [REDCap](http://www.project-redcap.org/) through its API.

## Reading REDCap Data
The function `redcap_read_oneshot` uses the [`RCurl`](http://cran.r-project.org/web/packages/RCurl/index.html) package to call the REDCap API.
```{r set_options, echo=FALSE, results='hide'}
require(knitr)
opts_chunk$set(
    comment=NA, 
    tidy=FALSE
)
# options(markdown.HTML.header = system.file("misc", "vignette.css", package = "knitr"))
# options(markdown.HTML.header = system.file("misc", "vignette.css", package = "REDCapR"))
# options(markdown.HTML.header = file.path(devtools::inst("REDCapR"), "misc", "vignette.css"))

options(width=120) #So the output is 50% wider than the default.
```

### Set project-wide values.
There is some information that is specific to the REDCap project, as opposed to an individual operation.  This includes the (1) uri of the server, the (2) token for the user's project, and the (3) location of the SSL certification store, which is necessary to verify the identity of the REDCap webserver.  

The `REDCapR` package includes a recent version of the [Bundle of CA Root Certificates](http://curl.haxx.se/ca/cacert.pem) from the official [cURL site](http://curl.haxx.se).  This version is used by default, unless the `cert_location` parameter is given another location.

```{r project_values}
library(REDCapR) #Load the package into the current R session.
uri <- "https://bbmc.ouhsc.edu/redcap/api/"
token <- "9A81268476645C4E5F03428B8AC3AA7B"
```

### Read all records and fields.
If no information is passed about the desired records or fields, then the entire data set is returned.  Only two parameters are required, `redcap_uri` and `token`.  Unless the `verbose` parameter is set to `FALSE`, a message will be printed on the R console with the number of records and fields returned.

```{r return_all}
#Return all records and all variables.
ds_all_rows_all_fields <- redcap_read_oneshot(redcap_uri=uri, token=token)$data
ds_all_rows_all_fields #Inspect the returned dataset
```

### Read a subset of the records.
If only a subset of the **records** is desired, the two approaches are shown below.  The first is to pass an array (where each element is an ID) to the `records` parameter.  The second is to pass a single string (where the elements are separated by commas) to the `records_collapsed` parameter.  

The first format is more natural for more R users.  The second format is what is expected by the REDCap API.  If a value for `records` is specified, but `records_collapsed` is missing, then `redcap_read_oneshot` automatically converts the array into the format needed by the API.

```{r read_row_subset, results='hold'}
#Return only records with IDs of 1 and 3
desired_records_v1 <- c(1, 3)
ds_some_rows_v1 <- redcap_read_oneshot(
   redcap_uri = uri, 
   token = token, 
   records = desired_records_v1
)$data

#Return only records with IDs of 1 and 3 (alternate way)
desired_records_v2 <- "1, 3"
ds_some_rows_v2 <- redcap_read_oneshot(
   redcap_uri = uri, 
   token = token, 
   records_collapsed = desired_records_v2
)$data

ds_some_rows_v2 #Inspect the returned dataset
```

### Read a subset of the fields.
If only a subset of the **fields** is desired, then two approaches exist.  The first is to pass an array (where each element is an field) to the `fields` parameter.  The second is to pass a single string (where the elements are separated by commas) to the `fields_collapsed` parameter. Like with `records` and `records_collapsed` described above, this function converts the more natural format (*ie*, `fields`) to the format required by the API (*ie*, `fields_collapsed`) if `fields` is specified and `fields_collapsed` is not.

```{r read_field_subset}
#Return only the fields recordid, first_name, and age
desired_fields_v1 <- c("recordid", "first_name", "age")
ds_some_fields_v1 <- redcap_read_oneshot(
   redcap_uri = uri, 
   token = token, 
   fields = desired_fields_v1
)$data

#Return only the fields recordid, first_name, and age (alternate way)
desired_fields_v2 <- "recordid, first_name, age"
ds_some_fields_v2 <- redcap_read_oneshot(
   redcap_uri = uri, 
   token = token, 
   fields_collapsed = desired_fields_v2
)$data

ds_some_fields_v2 #Inspect the returned dataset
```

### Additional Returned Information
The examples above have shown only the resulting `data.frame`, by specifying `$data` at the end of the call.  However, more is available to those wanting additional information, such as
 1. The `data` object has the `data.frame`, as in the previous examples.
 1. The `success` boolean value indicates if `redcap_read_oneshot` believes the operation completed as intended.
 1. The `status_codes` is a collection of [http status codes](http://en.wikipedia.org/wiki/List_of_HTTP_status_codes), separated by semicolons.  There is one code for each batch attempted.
 1. The `outcome_messages`: A collection of human readable strings indicating the operations' semicolons.  There is one code for each batch attempted.  In an unsuccessful operation, it should contain diagnostic information.
 1. The `records_collapsed` field passed to the API.  This shows which record subsets, if any, were requested.
 1. The `fields_collapsed` fields passed to the API.  This shows which field subsets, if any, were requested.
 1. The `elapsed_seconds` measures the duration of the call. 

```{r read_not_just_dataframe}
#Return only the fields recordid, first_name, and age
all_information <- redcap_read_oneshot(
   redcap_uri = uri, 
   token = token, 
   fields = desired_fields_v1
)
all_information #Inspect the additional information
```

## Session Info
For the sake of documentation and reproducibility, the current vignette was build on a system using the following software.

```{r session_info, echo=FALSE}
cat("Report created by", Sys.info()["user"], "at", strftime(Sys.time(), "%F, %T %z"))
sessionInfo()
```
