---
title: "Typical REDCap Workflow for a Data Analyst"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Typical REDCap Workflow for a Data Analyst}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy    = FALSE
)
```

```{r setup}
library(REDCapR)
```

Pre-requisites
===================================

Verify REDCapR is installed
-------------------------

First confirm that the REDCapR package is installed on your local machine.
```{r pre-req}
requireNamespace("REDCapR")
```

If this call fails, follow the installation instructions on the [REDCapR](https://ouhscbbmc.github.io/REDCapR) webpage.


Verify REDCap Access
-------------------------

Talk to your institution's REDCap administrator to ensure you have

* access to the REDCap server
* access to the specific REDCap project
* an API token with appropriate privileges ot the specific REDCap project.

Review Codebook
-------------------------

Before developing any REDCap or analysis code, take 10+ minutes to discuss the project with the investigator and to read the codebook.  The link to open codebook is near the top left corner, under the "Project Home and Design" banner.  Spending time to learn the details and context of the dataset will save you time later and improve the quality of the research.


Retrieve Protected Token
===================================

The REDCap API token is essentially a combination of your personal password and the ID of the specific project you're requesting data from.  Protect it like any other password to a PHI (protected health information](https://www.hhs.gov/answers/hipaa/what-is-phi/index.html)) dataset.  For a project with PHI, **never hard-code the password directly in an R file**.

Instead, we suggest storing the token in a location that can be accessed only by you.  We have two recommendations.

Security Method 1: Token File
-------------------------

The basic goals are (a) separate the secret values from the R file into a dedicated file and (b) secure the dedicated file.

The [`retrieve_credential_local()`](https://ouhscbbmc.github.io/REDCapR/reference/retrieve_credential.html) function in the [REDCapR](https://ouhscbbmc.github.io/REDCapR) package reads relevant information from a csv and loads them into R.

The plain-text file might look like this:


```csv
redcap_uri,username,project_id,token,comment
"https://ou.edu/redcap/api/","myusername","153","9A81268476645C4E5F03428B8AC3AA7B","simple"
"https://ou.edu/redcap/api/","myusername","212","0434F0E9CF53ED0587847AB6E51DE762","longitudinal"
"https://ou.edu/redcap/api/","myusername","213","D70F9ACD1EDD6F151C6EA78683944E98","simple write data"
```

To retrieve the credentials for the first project listed above, pass the value of "153" to `project_id`.

```{r retrieve-credential}
path_credential <- system.file("misc/example.credentials", package = "REDCapR")
credential  <- REDCapR::retrieve_credential_local(
  path_credential = path_credential, 
  project_id      = 153
)

credential
```

Compared to the method below, this one is less secure but easier to establish.

Security Method 2: Token Server
-------------------------

Our preferred method involves saving the tokens in a separate database that uses something like Active Directory to authenticate requests.  This method is described in detail in the [Security Database](https://ouhscbbmc.github.io/REDCapR/articles/SecurityDatabase.html) vignette.  

This approach realistically requires someone in your institution to have at least basic database administration experience. 

Read Data: Unstructured Approach 
===================================


```{r unstructured}
ds_1 <- 
  REDCapR::redcap_read(
    redcap_uri  = credential$redcap_uri, 
    token       = credential$token
  )$data

ds_1
```

Read Data: Choosing Columns and Rows
===================================

When you read a dataset for the first time, you probably haven't decided which columns are needed so it makes sense to retrieve them all.  As you gain familiarity with the data and with the analytic objectives consider being more selective with the variables and rows transported from the remote server to your local machine.  

Performance advantages include:

1. A server is almost always more efficient filtering than a language like R or Python.
1. REDCap's PHP code retrieves less data from the database and translates less to a text format (like csv or json).
1. Fewer bytes are transmitted across your network.
1. Your local machine will have better performance, because R has a smaller dataset to manage.
1. Your brain doesn't have to look past unnecessary columns.
1. Your R code doesn't have filter what the server already removed.
1. Highly-sensitive PHI columns that are unnecessary for an analysis remain on the server.

Read Data: Structured Approach 
===================================



Next Steps
===================================

Create an Arch for reuse
-------------------------

Notes
===================================

This vignette was originally designed for a 2021 R/Medicine REDCap workshop with [Peter Higgins](https://research.medicine.umich.edu/department/staff/peter-higgins-md-phd), [Amanda Miller](https://coloradosph.cuanschutz.edu/resources/directory/directory-profile/Miller-Amanda-UCD6000053152), and [Kenneth McLean](https://twitter.com/kennethmclean92).
